% =====================================================================
\section{Related Work}
\label{sec:related}
% =====================================================================

\paragraph{Manifold-aware Shapley attribution.}
The KernelSHAP value function~\citep{lundberg2017unified} requires
imputing absent features.  In the original formulation, absent features are
drawn from the marginal distribution, producing \emph{off-manifold}
explanations that may evaluate the model on inputs never seen during
training~\citep{frye2021shapley}.  \citet{aas2021explaining} introduce three
on-manifold imputers based on Gaussian, Gaussian-copula, and empirical-conditional
approximations of the conditional distribution; this approach is extended to
non-parametric vine copulas~\citep{aas2021vineexplaining} and to mixed
data~\citep{redelmeier2020explaining}.  \citet{olsen2022using} model the
conditional distribution with a single variational autoencoder with arbitrary
conditioning (VAEAC)~\citep{ivanov2018variational}, avoiding the need to fit
a separate model for each coalition.  Most recently,
\citet{zhang2026axiomatic} use optimal-transport flow matching to produce
Aumann--Shapley-type attributions with on-manifold paths.
None of these methods is benchmarked on data with non-trivial temporal structure.

\paragraph{Temporal Shapley attribution.}
A separate line of work extends Shapley attribution to time-series data.
TimeSHAP~\citep{bento2021timeshap} reformulates KernelSHAP at the
time-step or window level, replacing absent prefixes with a baseline
sequence; we refer to this approach as \emph{KernelSHAP--Temporal} for
consistency with our naming convention.  WindowSHAP~\citep{nayebi2023windowshap}
extends this to overlapping sliding windows.
\citet{villani2022feature} replace the linear surrogate with a vector autoregression
to produce time-consistent Shapley values.  ShaTS~\citep{de2025shats}
introduces temporal/sensor/process-level player groupings.
\citet{kim2026groupsegment} construct group-segment players that respect
nonlinear feature dependencies and temporal distribution shifts.
All of these approaches use marginal or zero imputers and therefore retain the
off-manifold pathology of the original KernelSHAP formulation.

\paragraph{Shapley attribution for human motion.}
Direct applications of SHAP to human motion are sparse.
\citet{tempel2025choose, auletta2023predicting} apply standard SHAP to
gait classifiers without addressing the manifold or temporal structure.
\citet{tempel2025explaining} introduce ShapGCN, an architecture-specific
GCN edge-importance method that does not generalise to non-graph encoders.
No benchmark exists that systematically evaluates Shapley attribution on
structured spatiotemporal motion data, which is the gap our work addresses.

\paragraph{Benchmarks for explainability.}
Synthetic benchmarks with controllable ground-truth attributions
have been instrumental in evaluating explanation
methods~\citep{arras2022clevr, hedstrom2023quantus, Yang2019BenchmarkingAM}.
Quantus~\citep{hedstrom2023quantus} provides a comprehensive metric suite for
faithfulness, robustness, and complexity.  Our benchmark complements this
ecosystem with a parametric family of \emph{spatiotemporal} synthetic datasets
and a player-set abstraction that has not previously been benchmarked.
