% =====================================================================
\section{Background}
\label{sec:background}
% =====================================================================

\subsection{Shapley Values for Feature Attribution}

Let $f : \mathcal{X} \to \mathbb{R}$ be a scalar-valued predictor — concretely,
the predicted probability $f_y(\mathbf{x})$ that input $\mathbf{x} \in \mathcal{X}$
belongs to class $y$.  Given $n$ interpretable \emph{players}
$N = \{1, \ldots, n\}$ (e.g.\ anatomical joints or temporal windows), we seek
additive attributions $\phi_1, \ldots, \phi_n \in \mathbb{R}$ such that
\begin{equation}
  f_y(\mathbf{x}) \approx \phi_0 + \sum_{i=1}^{n} \phi_i,
  \label{eq:additive}
\end{equation}
where $\phi_0$ is a baseline reference value.

\paragraph{Coalitions and the value function.}
A \emph{coalition} $S \subseteq N$ identifies the players that are
\emph{present}; its complement $\bar{S} = N \setminus S$ contains the
\emph{absent} players.  Because $f$ requires a complete input, absent players
must be imputed.  Let $\mathbf{x}_S$ denote the components of $\mathbf{x}$
indexed by $S$, and let $\mathbf{x}'_{\bar{S}}$ be a draw for the absent
components.  Two natural choices of imputation distribution define the two
canonical value functions~\citep{frye2021shapley}:
\begin{align}
  v^{(\mathrm{off})}_\mathbf{x}(S)
    &= \mathbb{E}_{p(\mathbf{x}')}\!\left[
        f_y\!\left(\mathbf{x}_S \sqcup \mathbf{x}'_{\bar{S}}\right)
      \right],
  \label{eq:v_off} \\
  v^{(\mathrm{on})}_\mathbf{x}(S)
    &= \mathbb{E}_{p(\mathbf{x}' \mid \mathbf{x}_S)}\!\left[
        f_y\!\left(\mathbf{x}_S \sqcup \mathbf{x}'_{\bar{S}}\right)
      \right],
  \label{eq:v_on}
\end{align}
where $\mathbf{x}_S \sqcup \mathbf{x}'_{\bar{S}}$ denotes the input formed by
splicing observed components $\mathbf{x}_S$ with imputed components
$\mathbf{x}'_{\bar{S}}$.  Eq.~\eqref{eq:v_off} samples absent features
\emph{independently} of $\mathbf{x}_S$, so the resulting \emph{off-manifold}
Shapley values evaluate the model on potentially unrealistic spliced
inputs~\citep{frye2021shapley}.  Eq.~\eqref{eq:v_on} conditions absent features
on the observed coalition, keeping imputations on the data manifold.

\paragraph{Shapley values.}
Given a value function $v_\mathbf{x} : 2^N \to \mathbb{R}$, the
\emph{Shapley value} of player $i$ is~\citep{shapley1953value}
\begin{equation}
  \phi_i
  = \sum_{S \subseteq N \setminus \{i\}}
    \frac{|S|\,!\;(n - |S| - 1)!}{n!}
    \Bigl[v_\mathbf{x}(S \cup \{i\}) - v_\mathbf{x}(S)\Bigr].
  \label{eq:shapley}
\end{equation}
Shapley values are the unique attribution satisfying efficiency, symmetry,
linearity, and the dummy axiom~\citep{shapley1953value}.  Efficiency yields
the \emph{completeness} property
$\sum_{i=1}^n \phi_i = v_\mathbf{x}(N) - v_\mathbf{x}(\emptyset)$.

\subsection{Kernel SHAP}
\label{sec:kernelshap}

Exact evaluation of Eq.~\eqref{eq:shapley} requires $2^n$ queries of
$v_\mathbf{x}$.  \emph{Kernel SHAP}~\citep{lundberg2017unified} recovers
Shapley values as the solution to a weighted least-squares problem.
Working with the binary-vector representation $z' \in \{0,1\}^n$ (where
$z'_i = \mathbf{1}[i \in S]$), one fits the linear surrogate
\begin{equation}
  g(z') = \phi_0 + \sum_{i=1}^{n} \phi_i z'_i
  \label{eq:surrogate}
\end{equation}
by minimizing
\begin{equation}
  \hat{\boldsymbol{\phi}}
  = \arg\min_{\phi_0,\,\boldsymbol{\phi}}
    \sum_{z' \in \mathcal{Z}}
    \pi_n(z')\bigl[v_\mathbf{x}(z') - g(z')\bigr]^2,
  \label{eq:wls}
\end{equation}
where the Shapley kernel
$\pi_n(z') = \frac{n - 1}{\binom{n}{|z'|}\,|z'|\,(n - |z'|)}$
weights coalitions of size $|z'| \in \{1, \ldots, n-1\}$, and the boundary
coalitions $|z'| \in \{0, n\}$ are enforced as hard
constraints~\citep{lundberg2017unified}.
When $n$ is small (here $n \le 4$ temporal windows), all $2^n$ coalitions can
be enumerated exactly and Eq.~\eqref{eq:wls} is solved in closed form.

\paragraph{Role of the value function.}
The Kernel SHAP estimator is independent of how absent players are imputed.
Different choices instantiate different cooperative games and therefore
explain different questions: off-manifold value functions (Eq.~\ref{eq:v_off})
attribute model behaviour on potentially unrealistic inputs, whereas the
on-manifold value function (Eq.~\ref{eq:v_on}) attributes the model's
dependence on information genuinely present in the data~\citep{frye2021shapley}.
For motion sequences, where joints are biomechanically coupled and temporal
windows are phase-dependent, this distinction is consequential.

\subsection{Temporal SHAP Variants}
\label{sec:temporal-shap}

Two existing methods adapt Kernel SHAP to time-series data; both retain
the off-manifold formulation of Eq.~\eqref{eq:v_off}.
\textbf{KernelSHAP-Temporal} (TimeSHAP~\citep{bento2021timeshap})
defines players as ordered time-step prefixes (or windows), with the
value function replacing absent prefixes by draws from a baseline reference
sequence (typically all-zero or training mean).  Coalitions correspond to
contiguous prefixes of the sequence rather than arbitrary subsets,
inducing an ordering-aware coalition structure.
\textbf{WindowSHAP}~\citep{nayebi2023windowshap} partitions the sequence into
fixed-width sliding windows and applies Kernel SHAP at the window level,
again imputing absent windows from a baseline reference.
Both methods use temporal-window players, but neither imputes from a
data-conditional distribution; they cannot therefore distinguish whether
a player's contribution is genuine or arises from the off-manifold pathology
discussed above.  Our benchmark places these methods in a unified comparison
with on-manifold imputers.
