% ======================================================================
\section{Benchmark Construction}
\label{sec:method}
% ======================================================================

The benchmark is organised around three orthogonal design choices:
\emph{(i)}~the player-set abstraction (Sec.~\ref{sec:players});
\emph{(ii)}~the conditional imputer used by KernelSHAP
(Sec.~\ref{sec:attribution-methods}); and
\emph{(iii)}~the dataset-generation process and corresponding oracle
(Sec.~\ref{sec:synthetic} and~\ref{sec:real-data}).
We use the spatiotemporal motion notation
$\mathbf{x} \in \mathbb{R}^{J \times F \times T}$ throughout, with $J$
joints, $F$ Cartesian channels, and $T$ frames; for the real BMCLab data
we use the Human3.6M skeleton ($J{=}17$, $F{=}3$,
$T{=}80$~\citep{ionescu2014human36m}); synthetic datasets vary $J$ and
fix $T{=}16$.  A coalition $S \subseteq N$ over $n$ players induces a
mask $\mathbf{m}_S \in \{0,1\}^{J \times F \times T}$ that is $1$ on
observed coordinates and $0$ on absent coordinates; the imputed
coalition value follows Eq.~\eqref{eq:v_on}, evaluated by Monte-Carlo
average over $M$ completions:
\begin{equation}
  \hat{v}^{(\mathrm{on})}_{\mathbf{x}}(S)
  = \frac{1}{M}\sum_{r=1}^{M}
    f_y\!\left(
      \mathbf{m}_S \odot \mathbf{x}
      + (1 - \mathbf{m}_S) \odot \hat{\mathbf{x}}^{(r)}
    \right),
  \quad
  \hat{\mathbf{x}}^{(r)} \sim p_\theta(\cdot \mid \mathbf{x}\odot\mathbf{m}_S, \mathbf{m}_S).
  \label{eq:vfn_mc}
\end{equation}

%-----------------------------------------------------------------------
\subsection{Spatiotemporal Player Sets}
\label{sec:players}
%-----------------------------------------------------------------------

A key design choice in Shapley attribution for structured data is the
\emph{player set}: which subsets of $\{1,\dots,J\}\!\times\!\{1,\dots,F\}\!\times\!\{1,\dots,T\}$
constitute a single ``player'' whose removal is a unit coalitional step.
Finer granularity gives more detailed attributions but exponential
coalition cost; coarser groupings are tractable but may blend signal from
multiple features.  Our \texttt{PlayerSet} API exposes three canonical
abstractions, all interoperable with every attribution method we benchmark.

\textbf{Temporal windows} $\mathcal{P}_{\mathrm{temp}}$ partition the
$T$ frames into $K$ contiguous windows (uniform on synthetic data;
gait-phase windows on real data, with phase boundaries from the
autocorrelation of the ankle--pelvis trajectory at $\varphi
\in (0.30,\,0.60,\,0.80)$); $K{\in}\{4,8\}$ yielding
$2^K{\in}\{16,256\}$ coalitions evaluated exactly.
\textbf{Spatial joints} $\mathcal{P}_{\mathrm{joint}}$ have $n=J$
players each spanning one joint across all $F$ channels and $T$ frames;
anatomical-group attributions are obtained by additive aggregation
(efficiency-preserving).
\textbf{Joint-window cells} $\mathcal{P}_{\mathrm{cell}}$ have
$n = J\times K$ players (used in supplementary cross-granularity
diagnostics).

%-----------------------------------------------------------------------
\subsection{Attribution Methods}
\label{sec:attribution-methods}
%-----------------------------------------------------------------------

\paragraph{KernelSHAP imputer variants.}
Six methods share the same player set ($\mathcal{P}_{\mathrm{temp}}$,
$K{=}4$), the same Kernel SHAP weighted-least-squares estimator
(Sec.~\ref{sec:kernelshap}), and the same exact 16-coalition
enumeration.  They differ \emph{only} in the conditional distribution
$p_\theta(\cdot \mid \mathbf{x}\odot\mathbf{m}_S, \mathbf{m}_S)$ used in
Eq.~\eqref{eq:vfn_mc}.
\textbf{KS--Zero}: uniform off-manifold baseline
($\hat{\mathbf{x}}^{(r)}{=}\mathbf{0}$) from the original KernelSHAP
formulation~\citep{lundberg2017unified}.
\textbf{KS--Mean}: uniform off-manifold baseline using the per-channel
training mean~\citep{lundberg2017unified}.
\textbf{KS--Marginal}: the marginal-product imputer of
\citet{aas2021explaining}; each absent player's coordinates drawn from a
random training donor, respecting marginal statistics but ignoring
$\mathbf{x}_S$.
\textbf{KS--Empirical}: the empirical-conditional imputer of
\citet{aas2021explaining,redelmeier2020explaining}; conditional
kernel-weighted donor sampling (approximately on-manifold).
\textbf{KS--VAEAC}: the variational on-manifold imputer of
\citet{olsen2022using}, implemented via VAE-with-arbitrary-conditioning
\citep{ivanov2018variational}.
\textbf{KS--Flow}: a flow-matching analogue inspired by
\citet{zhang2026axiomatic}, with RePaint harmonisation
\citep{lugmayr2022repaint}, integrated by Euler with $20$ steps
(on-manifold).

\paragraph{Temporal SHAP variants.}
We additionally evaluate KernelSHAP-Temporal~\citep{bento2021timeshap}
and WindowSHAP~\citep{nayebi2023windowshap}, which are described in
Sec.~\ref{sec:temporal-shap}.  Both use temporal-prefix or sliding-window
players with off-manifold (zero or mean) reference imputation, so any
performance gap with respect to KS--Mean or KS--Zero attributable to the
\emph{coalition-structure} (rather than to manifold awareness) can be
isolated.

\paragraph{Gradient-based baselines.}
Integrated Gradients~\citep{sundararajan2017axiomatic},
DeepLIFT~\citep{shrikumar2017learning},
SmoothGrad~\citep{smilkov2017smoothgrad}, and
LRP~\citep{bach2015pixel} form the gradient-based baselines.  These
methods do not invoke a coalition value function and therefore do not
benchmark imputation quality; we include them in the main EC1 table
(Table~\ref{tab:synth_ec1}) to make the oracle-anchored cost of
choosing a non-Shapley attribution explicit, addressing the broader
question of whether one should use Shapley attribution at all on
spatiotemporal data.  All four are aggregated to player-level scores by
summing per-coordinate $|\nabla_x f_y|$ within each player.  LRP fails
on transformer architectures due to known propagation issues with
self-attention~\citep{ali2022xai}; those cells are excluded.

%-----------------------------------------------------------------------
\subsection{On-Manifold Imputers (Implementation)}
\label{sec:on-manifold-imputers}
%-----------------------------------------------------------------------

KS--VAEAC and KS--Flow share a transformer architecture from the
CARE-PD codebase~\citep{adeli2025multi} trained with mixed temporal-window,
spatial-joint, and element-wise coalition masks: $400$ epochs on BMCLab
($T{=}80$, $J{=}17$, $F{=}3$) for the real-data sweep, $80$ epochs
re-trained per generative process for synthetic.  Both expose a
\texttt{sample\_completions\_batched} interface that generates all $2^K$
coalition completions in a single GPU pass.  KS--VAEAC samples from the
conditional prior and decodes; KS--Flow integrates the conditional ODE
from $\mathbf{x}_0\!\sim\!\mathcal{N}(\mathbf{0},\mathbf{I})$ to $t{=}1$
with $20$ Euler steps and RePaint harmonisation~\citep{lugmayr2022repaint}.
Observed entries are bit-for-bit overwritten before evaluating $f_y$.

%-----------------------------------------------------------------------
\subsection{Synthetic Dataset Suite}
\label{sec:synthetic}
%-----------------------------------------------------------------------

We introduce a parametric family of synthetic motion datasets for which
ground-truth Shapley values are available either in closed form or by
exact enumeration of a tractable conditional distribution.  All datasets
share shape $\mathbb{R}^{J \times F \times T}$ with $T{=}16$, an
\emph{identical} label-generation procedure (described below), and
support both $\mathcal{P}_{\mathrm{temp}}$ and $\mathcal{P}_{\mathrm{joint}}$
player sets.  Table~\ref{tab:datasets} summarises the family.

\input{tables/table1_dataset_taxonomy}

\paragraph{Unified covariance structure.}
The latent Gaussian field $\mathbf{z}$ underlying every dataset has
Kronecker-product covariance
\begin{equation}
  \operatorname{Cov}(\mathrm{vec}\,\mathbf{z})
  = \Sigma_J \otimes \Sigma_F \otimes \Sigma_T,
  \label{eq:kronecker}
\end{equation}
with $\Sigma_F = I_F$ throughout, while $\Sigma_J$ and $\Sigma_T$ vary
across datasets to systematically span the (low/high spatial) $\times$
(low/high temporal) design grid.  The Burr-marginal datasets
(\texttt{burr\_m5}, \texttt{burr\_m10}) apply a heavy-tailed copula
transform $x_{jft} = \operatorname{sign}(z_{jft})Q_{\mathrm{Burr}}(2\Phi(|z_{jft}|)-1; c, k)$
with $(c,k){=}(2,2)$, retaining the latent covariance structure but giving
tail index $ck{=}4$.

\paragraph{Label and classifier.}
For each dataset we compute window-mean features
$w_k(\mathbf{x}) = \mathbb{E}_{j,f,t \in W_k}[x_{jft}]$ and apply
the Olsen interaction~\citep{olsen2022using}:
\begin{equation}
  s(\mathbf{x})
  = \sum_r \Bigl[
    a_r \sin(\pi u_{4r} u_{4r+1})
    + b_r u_{4r+2} \exp(c_r u_{4r+2} u_{4r+3})
  \Bigr],
  \quad u_k = \Phi(w_k/\sigma_k),
  \label{eq:olsen}
\end{equation}
with random coefficients $(a_r, b_r, c_r)$ per block of four players.
Class labels are tertiles of $s(\mathbf{x})$ on a calibration set.
The classifier is a small MLP, CNN, or Transformer trained on the
window-mean features, all reported separately.

\paragraph{Oracle.}
For Gaussian datasets, the conditional distribution
$p(\mathbf{z}_{\bar S}\mid\mathbf{z}_S)$ is available in closed form, so
the value function
$v(\mathbf{x};S) = \mathbb{E}[f_y(\mathbf{x}_S\sqcup\mathbf{z}'_{\bar S})\mid \mathbf{z}_S]$
can be evaluated exactly by Monte-Carlo over conditional Gaussian draws.
Combined with exact enumeration of the $2^K$ coalitions, this gives
ground-truth Shapley values to within Monte-Carlo noise.  Burr datasets
use the same procedure after the analytic copula inverse mapping.  The
combined high-spatial $\times$ strong-temporal regime
(\texttt{skeleton\_gait\_combined}) uses the same Kronecker-product
covariance with $\Sigma_J$ taken from \texttt{skeleton} and $\Sigma_T$
from \texttt{gait\_periodic}; because Kronecker products preserve closed-form
conditionals, the oracle is identical to the single-axis case --- only
the matrices differ.

%-----------------------------------------------------------------------
\subsection{Real-World Dataset: BMCLab Gait (CARE-PD)}
\label{sec:real-data}
%-----------------------------------------------------------------------

For real-world evaluation we use BMCLab from the CARE-PD
study~\citep{adeli2025multi}: 80-frame H36M clips at 27\,fps with UPDRS-gait
severity labels in $\{0, 1, 2\}$.  We evaluate three leave-one-subject-out
folds with $n{=}200$ clips per fold, using fine-tuned
MotionBERT~\citep{zhu2023motionbert} and $K{=}4$ gait-phase windows.
Ground-truth Shapley values are unavailable; we therefore report
faithfulness correlation \emph{and} PlayerAOPC, with disagreement
between the two interpreted as imputer-realism evidence
(Sec.~\ref{sec:results-real}).

%-----------------------------------------------------------------------
\subsection{Evaluation Metrics}
\label{sec:metrics}
%-----------------------------------------------------------------------

We adopt the on-manifold Shapley evaluation framework of
\citet{olsen2022using} and extend it from tabular to spatiotemporal data.

\paragraph{Attribution error (EC1, EC2, EC3 — synthetic only).}
Given estimated attributions $\hat{\boldsymbol\phi}\in\mathbb{R}^n$ and
oracle Shapley values $\boldsymbol\phi^*\in\mathbb{R}^n$:

\textbf{EC1 (mean absolute Shapley error)} measures the average
absolute deviation between estimated and oracle per-player Shapley
values, in units of prediction probability.  Introduced as the primary
criterion in the on-manifold KernelSHAP literature by
\citet{aas2021explaining} and \citet{olsen2022using}, and equivalent to
the ``mean absolute Shapley deviation'' of \citet{frye2021shapley}; lower
is better:
$\mathrm{EC1} = \frac{1}{n}\|\hat{\boldsymbol\phi} - \boldsymbol\phi^*\|_1$.

\textbf{EC2 (mean squared Shapley error)} is the average squared
deviation, more sensitive to large per-player misattributions
\citep{olsen2022using}; lower is better:
$\mathrm{EC2} = \frac{1}{n}\|\hat{\boldsymbol\phi} - \boldsymbol\phi^*\|_2^2$.

\textbf{EC3 (Shapley rank decorrelation)} isolates ranking errors from
magnitude errors: $\mathrm{EC3} = 1 - \rho_\text{Pearson}(\hat{\boldsymbol\phi},
\boldsymbol\phi^*)$ \citep{olsen2022using}; lower is better, 0 means
perfect rank agreement.

\textbf{Top-$k$ recovery} measures the fraction of the $k$ highest-magnitude
oracle players that the estimator also ranks in its top-$k$
\citep{olsen2022using,hooker2019roar}; Spearman and Kendall rank
correlations are reported in supplementary.

\paragraph{Faithfulness (synthetic and real).}
\textbf{Faithfulness correlation}~\citep{bhatt2020evaluating,hedstrom2023quantus}
is the Pearson correlation between $\sum_{i\notin S}\phi_i$ and
$v(N){-}v(S)$ across all coalitions; higher means attribution magnitudes
track the model's actual coalition-removal sensitivity.
\textbf{PlayerAOPC} (deletion-curve area) is the mean prediction drop
when players are removed in decreasing order of $|\phi|$, following the
ERASER-style comprehensiveness criterion~\citep{deyoung2020eraser};
higher means more concentrated, falsifiable attributions.  Both metrics
reuse the $v(S)$ values KernelSHAP already evaluates, with no additional
classifier calls.

