% =====================================================================
\section{Synthetic Benchmark Results}
\label{sec:results-synth}
% =====================================================================

We organise the synthetic results along two axes: spatial and temporal
complexity.  We report attribution error EC1 against oracle Shapley values
(Sec.~\ref{sec:results-error}), the 2$\times$2 spatiotemporal failure-mode
matrix (Sec.~\ref{sec:results-2x2}), and the faithfulness and ranking
metrics that transfer to the real-data setting
(Sec.~\ref{sec:results-faithfulness}).

\subsection{Attribution Error vs.\ Oracle Shapley Values}
\label{sec:results-error}

Table~\ref{tab:synth_ec1} reports EC1 across five pillar datasets spanning
the 2$\times$2 spatial $\times$ temporal design and the heavy-tailed Burr
regime, averaged over three classifier architectures (MLP, CNN,
Transformer).  Four patterns are consistent across architectures:
\textbf{(i)} \emph{Learned conditional imputers recover oracle Shapley
values to within Monte Carlo noise.}  KS--VAEAC achieves EC1 $\le 0.007$
on every pillar dataset, including the high-spatial $\times$
strong-temporal regime (\texttt{skel+gait}, EC1 $0.007$) and the
heavy-tailed \texttt{burr\_m5} (EC1 $0.004$).  KS--Flow tracks within a
factor of two and narrowly wins on \texttt{gait\_periodic}.
The conditional Gaussian oracle row (Table~\ref{tab:synth_ec1}) shows
residual MC noise: the oracle \emph{method} uses $n_\text{mc}{=}10$
random conditional samples per coalition (the same inference budget as
KernelSHAP imputers), and its EC1 of $0.06$--$0.12$ represents the
variance floor of this low-sample estimator.  Learned imputers
(KS--VAEAC, KS--Flow) substantially undercut the oracle method row
because each VAEAC sample approximates the conditional mean far more
accurately than a single Gaussian draw: effectively, the generator
``pre-conditions'' all relevant structure, making $M{=}1$ completion
nearly as good as $M{=}n_\text{mc}$.  Scaling to $n_\text{mc}{=}200$ is
predicted to reduce oracle EC1 by $\sqrt{10}\approx 3.16\times$,
converging toward the oracle's noiseless limit.  The learned imputers
already lie \emph{below} this limit at all tested datasets.
\textbf{(ii)} \emph{Off-manifold imputers cluster around a single error
level} (KS--Zero / KS--Mean / KS--Marginal $\in [0.08, 0.25]$), with no
clear winner; KS--Empirical (kernel-weighted donor sampling) is
competitive only on \texttt{gait\_periodic}.
\textbf{(iii)} \emph{Temporal-window players alone do not improve
attribution.}  KS--Temporal tracks KS--Zero closely; WindowSHAP performs
worst (EC1 $4$--$6\times$ the next-worst KernelSHAP variant).
WindowSHAP's failure is structural and not fixable by tuning window size:
Table~\ref{tab:winshap_diag} (Appendix) shows EC1 increasing monotonically
from $0.24$ ($w{=}2$) to $2.6$ ($w{=}16$) on \texttt{gauss\_k4}, and
from $0.67$ ($w{=}2$) to $1.6$ ($w{=}16$) on \texttt{skel+gait} ---
always $10$--$100\times$ above KS--VAEAC.  The root cause is that
WindowSHAP's sliding window-mean baseline assigns attribution mass to
boundary coalitions rather than interior frames, violating the
boundary-coalition constraint that KernelSHAP's weighted-least-squares
enforces, independent of $w$.
Temporal-aware coalition structure without manifold-aware imputation
offers no measurable benefit.
\textbf{(iv)} \emph{Gradient baselines (IG, DeepLIFT, SmoothGrad, LRP) are
not competitive on EC1}, with errors typically $5$--$50\times$ above
off-manifold KernelSHAP.  This is expected (gradient methods estimate
sensitivity, not cooperative-game value), but makes the oracle-anchored
cost of choosing a non-Shapley attribution explicit.

\paragraph{Player-set granularity.}
Comparing \texttt{gauss\_k4} ($K{=}4$) and \texttt{gauss\_k8} ($K{=}8$)
isolates the effect of player granularity: the on-manifold dominance is
unchanged, although $K{=}8$ doubles the number of coalitions ($2^{8}
{=}256$).  Table~\ref{tab:winners} (rightmost column) reports the best
EC1 across player-set granularities
($\mathcal{P}_{\mathrm{temp}}$, $\mathcal{P}_{\mathrm{joint}}$,
$\mathcal{P}_{\mathrm{cell}}$) for KS--VAEAC on the two high-spatial
regimes.  For \texttt{skeleton} (high spatial, AR(1) temporal),
$\mathcal{P}_{\mathrm{joint}}$ (17 spatial players) yields EC1 $\approx
0.010$--$0.020$ across classifiers, compared to $\mathcal{P}_{\mathrm{temp}}$
(4 temporal players) at $0.004$; the coarser temporal partition wins
because KernelSHAP's coalition-sampling budget covers 4 players far
better than 17, and VAEAC's generator already captures spatial structure
implicitly.  For \texttt{skel+gait} (high spatial, strong temporal), a
similar effect holds for $\mathcal{P}_{\mathrm{cell}}$ (68
joint$\times$window players).  In both regimes, on-manifold EC1 stays
below $0.007$ for $\mathcal{P}_{\mathrm{temp}}$, confirming that
on-manifold imputation benefit is robust to player-set choice and that
granularity should be matched to interpretability needs rather than
optimised for EC1.

\input{tables/table2_synth_ec1}

\subsection{The Spatial--Temporal Failure-Mode Matrix}
\label{sec:results-2x2}

\input{tables/table3_2x2_winners}

Table~\ref{tab:winners} quantifies the on-manifold benefit per regime:
the best on-manifold method beats the best off-manifold KernelSHAP
variant by $15$--$31\times$ on every quadrant tested, including the
high-spatial $\times$ strong-temporal regime (\texttt{skel+gait},
$25\times$) that most closely resembles real motion data.  KS--VAEAC
dominates three of the four regimes; KS--Flow narrowly dominates the
low-spatial strongly-temporal periodic regime, consistent with its
continuous-time formulation more naturally capturing periodic structure.
The off-manifold winner is regime-dependent (KS--Temporal in the two
AR(1) regimes, KS--Marginal in the two strongly-temporal regimes), but
absolute EC1 remains an order of magnitude above the on-manifold methods
regardless of which off-manifold method is chosen.
Heavy-tailed Burr-marginal data (\texttt{burr\_m5}, \texttt{burr\_m10})
is reported in Table~\ref{tab:synth_ec1}: we retrained VAEAC and Flow
imputers from scratch on the Burr-marginal distribution using the same
transformer architecture ($80$ epochs; see \texttt{REPRODUCIBILITY.md}).
KS--VAEAC and KS--Flow both reach EC1 $\le 0.005$ on \texttt{burr\_m5}
and \texttt{burr\_m10} despite tail index $ck{=}4$, matching their
Gaussian-pillar performance and confirming that on-manifold dominance is
not an artefact of light tails.

\subsection{Faithfulness and Ranking Diagnostics}
\label{sec:results-faithfulness}

Attribution-error metrics require ground-truth Shapley values and therefore
cannot be evaluated on real data.  We complement EC1 with two metrics that
\emph{do} transfer: faithfulness correlation and PlayerAOPC.  Both are
computed from the same coalition value vector $v(S)$ that KernelSHAP
already evaluates, so the additional cost is negligible.  On synthetic
data, faithfulness is moderately rank-correlated with EC1
(Spearman $\rho \approx 0.6$, supplementary), establishing that these
metrics are reasonable proxies in the absence of an oracle, but only as
\emph{relative} comparisons \emph{within a single value-function family}.

We highlight one caveat that motivates the multi-metric protocol of
Sec.~\ref{sec:results-real}: faithfulness is computed by removing players
in the order suggested by their Shapley values.  When the imputer produces
\emph{out-of-distribution} completions, removal causes large prediction
shifts that are easy to predict from any reasonable attribution, inflating
faithfulness without improving attribution accuracy.  This pathology is
most visible on real data where ground-truth Shapley values are unavailable.
The italic rows at the bottom of Table~\ref{tab:synth_ec1} show KS--VAEAC
EC1 as $M$ is varied from $1$ to $50$ on \texttt{gauss\_k4}: EC1 plateaus
by $M{=}5$ (gain $<0.001$ beyond that), justifying $M{=}1$ as the
main-sweep default while confirming the result is not an artefact of
underdrawing the conditional.
