% =====================================================================
\section{Real-World Application: CARE-PD Gait Classification}
\label{sec:results-real}
% =====================================================================

We now apply the benchmark to BMCLab gait sequences from the CARE-PD study
of Parkinson's-disease severity classification~\citep{adeli2025multi}.  The
classifier is the CARE-PD-fine-tuned MotionBERT~\citep{zhu2023motionbert};
inputs are 80-frame H36M-skeleton clips at 27\,fps; labels are UPDRS-gait
severity levels $\{0, 1, 2\}$.  We use $K{=}4$ \emph{gait-phase} windows
(loading, mid-stance, pre-swing, swing).  All five core KernelSHAP imputers
(KS--Zero, KS--Mean, KS--Marginal, KS--VAEAC, KS--Flow) are evaluated under
\emph{three} leave-one-subject-out folds with $n{=}200$ validation clips
per fold (600 total), and we report bootstrap 95\% confidence intervals
($B{=}10{,}000$ resamples) and a paired-bootstrap test for the
KS--Marginal versus KS--VAEAC faithfulness comparison.

\paragraph{The faithfulness--realism gap.}
Table~\ref{tab:real} reports faithfulness correlation and PlayerAOPC
for each imputer.  The off-manifold KS--Marginal achieves the highest
faithfulness ($+0.80$), substantially higher than the on-manifold
KS--VAEAC ($+0.53$) or KS--Flow ($+0.63$).
At first glance this contradicts the synthetic results
(Sec.~\ref{sec:results-synth}), where on-manifold methods dominated.
The contradiction is, however, a textbook example of a known pathology
in deletion-style faithfulness
metrics~\citep{hooker2019roar, jain2019attention}: when the imputer
produces out-of-distribution inputs, masking causes large prediction
shifts that are easy to predict from any reasonable attribution, inflating
the rank correlation.  The marginal-donor sequences place the classifier
in a regime far from training data --- producing the largest, easiest-to-rank
prediction drops --- without being more \emph{accurate} as Shapley values.

\input{tables/table4_real_carepd}

PlayerAOPC, which measures the mean drop magnitude rather than its
predictability, ranks methods in nearly the opposite order.  We do not
take this to mean that one of the two metrics is wrong; rather, both
report something genuine about the explanation that is conflated when only
one is reported.  We argue that benchmarking on real data where the
oracle is unavailable should always report \emph{both} a deletion-curve
metric (faithfulness or PlayerAOPC) \emph{and} a measure of imputer realism.
Our supplementary materials include qualitative comparisons of imputed
sequences for each method.

\paragraph{Per-phase attributions.}
On-manifold methods place positive attribution on the swing phase ---
the gait phase known to be most abnormal in Parkinsonian gait~\citep{morris1994ability}
--- with smaller absolute magnitude than off-manifold methods.  Off-manifold
methods produce larger-magnitude attributions but with phase rankings that
vary substantially across sequences, consistent with their imputations
introducing classifier-confidence shifts that dominate phase-specific
effects.  Per-phase Shapley vectors are reported in the supplementary.
