\documentclass[10pt]{article}

% NeurIPS 2026 — official style file should replace this when available.
\usepackage[margin=1in]{geometry}
\pdfmapfile{}  % avoid pdftex font-map warning that aborts pdflatex

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{nicefrac}
% \usepackage{microtype}  % disabled because system pdftex lacks scalable fonts (font expansion)
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{natbib}

\title{MotionBench-XAI: A Benchmark for Manifold-Aware\\
       Shapley Attribution on Temporal Data}

\author{%
  Anonymous Authors\\
  \texttt{anonymous@example.com}
}

\begin{document}

\maketitle

\begin{abstract}
Shapley-value attribution is often used for explaining machine learning models
on structured temporal data like human motion.  While well studied for temporal
data or manifold-structured data, Shapley explanations are not systematically
benchmarked for cases that require both.  We introduce MotionBench-XAI, a
benchmark for Shapley attribution on spatiotemporal data with an eye towards
gait classification as an example task.  We contribute (1) a parametric family
of synthetic temporal datasets with exact closed-form ground-truth Shapley
values, spanning Gaussian and heavy-tailed copula distributions and
spatiotemporal covariance structures, and (2) a unified evaluation platform
with spatiotemporal player-set abstractions, on- and off-manifold fidelity
metrics, attribution error, and cross-protocol ranking diagnostics.  We
benchmark sixteen attribution methods, including six KernelSHAP variants
differing only in imputer, on the synthetic suite and on a real-world
skeleton-based gait classification applied to Parkinson's disease severity
assessment.  All datasets, checkpoints, and code are released with a public
leaderboard.
\end{abstract}

\input{01-introduction}
\input{02-related}
\input{03-background}
\input{04-methods}
\input{05-results}
\input{06-results-real}

\section{Discussion}
\label{sec:discussion}

Two lessons emerge from our benchmark.
\emph{(i)} Learned conditional imputers (KS--VAEAC, KS--Flow) recover
oracle Shapley values to within Monte Carlo noise across every regime
tested --- including the high-spatial $\times$ strong-temporal regime
closest to real motion and the heavy-tailed Burr regime --- beating
every off-manifold or temporal-only competitor by an order of magnitude.
\emph{(ii)} Deletion-style faithfulness metrics reward imputers that
produce out-of-distribution completions; on CARE-PD the off-manifold
KS--Marginal obtains the highest faithfulness but the lowest PlayerAOPC,
with a significant paired-bootstrap gap (Sec.~\ref{sec:results-real}).
On real data we therefore recommend reporting any deletion-curve metric
alongside an imputer-realism diagnostic.

\paragraph{Limitations and reproducibility.}
The synthetic suite uses Kronecker-product covariance; real-world
evaluation is restricted to BMCLab gait + MotionBERT (3
leave-one-subject-out folds); generalisation to other motion
benchmarks and non-motion modalities is future leaderboard work.  All
data, checkpoints, and code are open-source at
\texttt{<anonymised url>}: \texttt{reproduce\_synthetic.sh} regenerates
Tables~\ref{tab:datasets}--\ref{tab:winners} in $\approx 40$\,min on one
GPU; \texttt{reproduce\_real.sh} regenerates Table~\ref{tab:real} given
data access (full instructions in \texttt{REPRODUCIBILITY.md}).

\bibliographystyle{plainnat}
\bibliography{references}  % uses references.bib (not references (1).bib)

% =====================================================================
\appendix
\section*{Supplementary Material}
% =====================================================================

\subsection*{WindowSHAP Window-Size Diagnostic}
\label{app:winshap}

\input{tables/table_winshap_diag}

Table~\ref{tab:winshap_diag} shows WindowSHAP EC1 as the window size $w$
is varied on the transformer architecture.  EC1 increases monotonically
with $w$, reaching $2.6\times$ at $w{=}16$ (the entire sequence).
This rules out a simple ``window too small/large'' explanation for
WindowSHAP's failure; rather, the root cause is structural:
WindowSHAP's sliding window-mean baseline assigns attribution mass to
the window boundary coalitions rather than interior frames, violating
the boundary-coalition constraint that KernelSHAP's weighted-least-squares
enforces.  Magnitude-shifted attributions result regardless of window size,
explaining why WindowSHAP performs worst across all regimes in Table~2.

\subsection*{$M$-Ablation (Completion Samples)}
\label{app:m_ablation}

EC1 for KS--VAEAC on \texttt{gauss\_k4} at $M\in\{1,5,20,50\}$
completion samples is reported as italic rows at the bottom of
Table~\ref{tab:synth_ec1} in the main text.  The result plateaus by
$M{=}5$, confirming $M{=}1$ as the main-sweep default.

\end{document}
